Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-21 talk/iwtc11/paper.tex
english tweeks.

XXX we either need to explain that we use C++ or consistently use C
\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
\hline
%NoBorder & 537.31 & 329.98 & 2.22 $\pm$ 0.03 & 2.17 $\pm$ 0.02 & - & -\\
%\hline
%NoBorder(iter) & 548.32 & 304.13 & 1.45 $\pm$ 0.03 & 1.47 $\pm$ 0.02 & - & -\\
%\hline
%NoBorder(range) & 534.64 & 317.34 & 1.34 $\pm$ 0.03 & 1.40 $\pm$ 0.04 & - & -\\
%\hline
conv3x3(NoBorderPadded(1000x1000)) & 543.73 & 333.20 & 2.09 $\pm$ 0.12 & 1.93 $\pm$ 0.05 & - & -\\
\hline
%NoBorderPadded(iter) & 546.70 & 309.32 & 1.21 $\pm$ 0.02 & 0.49 $\pm$ 0.02 & - & -\\
%\hline
%NoBorderPadded(range) & 550.92 & 318.33 & 1.12 $\pm$ 0.03 & 0.48 $\pm$ 0.01 & - & -\\
%\hline
conv3(array(1e5)) & 77.89 & 9.52 & 1.77 $\pm$ 0.06 & 0.68 $\pm$ 0.02 & 0.70 $\pm$ 0.05 & 0.59 $\pm$ 0.01\\
\hline
conv3(array(1e6)) & 77.15 & 9.58 & 1.69 $\pm$ 0.01 & 0.77 $\pm$ 0.01 & 0.84 $\pm$ 0.05 & 0.74 $\pm$ 0.01\\
\hline
conv3x3(Array2D(1000x1000)) & 23.72 & 12.77 & 0.07 $\pm$ 0.00 & 0.05 $\pm$ 0.03 & 0.24 $\pm$ 0.00 & 0.25 $\pm$ 0.01\\
\hline
conv3x3(Array2D(1000000x3)) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
\hline
conv5(array(1e5)) & 122.54 & 16.67 & 1.86 $\pm$ 0.02 & 1.05 $\pm$ 0.03 & 1.03 $\pm$ 0.05 & 0.65 $\pm$ 0.01\\
\hline
conv5(array(1e6)) & 125.77 & 16.80 & 1.92 $\pm$ 0.03 & 1.09 $\pm$ 0.02 & 1.07 $\pm$ 0.05 & 0.80 $\pm$ 0.01\\
\hline
dilate3x3(Array2D(1000x1000)) & 23.29 & 12.99 & 0.41 $\pm$ 0.04 & 0.39 $\pm$ 0.01 & 0.26 $\pm$ 0.00 & 0.26 $\pm$ 0.01\\
\hline
sobel(NoBorderPadded(1000x1000)) & 461.14 & 258.02 & 1.01 $\pm$ 0.03 & 0.48 $\pm$ 0.03 & - & -\\
\hline
sobel(Array2D(1000x1000)) & - & - & - & - & 0.19 $\pm$ 0.01 & 0.20 $\pm$ 0.01\\
\hline
%sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 $\pm$ 0.01 & 0.51 $\pm$ 0.00 & - & -\\
%\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93 $\pm$ 0.11 & 2.14 $\pm$ 0.02 & 0.97 $\pm$ 0.02 & 0.96 $\pm$ 0.01\\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 $\pm$ 0.03 & 1.00 $\pm$ 0.00 & 0.98 $\pm$ 0.00 & 0.98 $\pm$ 0.00\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 $\pm$ 0.01 & 1.82 $\pm$ 0.01 & 0.81 $\pm$ 0.01 & 0.80 $\pm$ 0.01\\
\hline

@@

The sobel and conv3x3 benchmarks are implemented
on top of two different two-dimensional array classes: Array2D
and NoBorderPadded. Array2D is 
a simple straight forward implementation providing 2 dimensionall
indexing with out of bounds checks. NoBorderPadded is optimized for
this specific application.
It ensures that there will be no failing guards, and makes
a lot of the two dimension index calculations loop invariant. The
intention here is twofold. It shows that the performance-impact of
having wrapper classes giving objects some application-specific
properties can be negligible. This is due to the inlining performed
during the tracing and the allocation removal of the index objects
introduced. It also shows that it is possible to do some low-level
hand optimizations of the Python code and hide those optimization
under a nice interface without loosing performance.

@@

The current approach still has some limitations which we plan to lift in the
future. In particular loop peeling is working less well in combination with
trace trees or trace stitching. The side exits attached guards that fail often
currently have to jump to the preamble which makes loops with several equally
common paths less efficient than they could be.

>>>>

XXX we either need to explain that we use C++ or consistently use C
\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
\hline
conv3(array(1e5)) & 77.89 & 9.52 & 1.77 $\pm$ 0.06 & 0.68 $\pm$ 0.02 & 0.70 $\pm$ 0.05 & 0.59 $\pm$ 0.01\\
\hline
conv3(array(1e6)) & 77.15 & 9.58 & 1.69 $\pm$ 0.01 & 0.77 $\pm$ 0.01 & 0.84 $\pm$ 0.05 & 0.74 $\pm$ 0.01\\
\hline
conv3x3(Array2D(1000x1000)) & 23.72 & 12.77 & 0.07 $\pm$ 0.00 & 0.05 $\pm$ 0.03 & 0.24 $\pm$ 0.00 & 0.25 $\pm$ 0.01\\
\hline
conv3x3(Array2D(1000000x3)) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
\hline
conv5(array(1e5)) & 122.54 & 16.67 & 1.86 $\pm$ 0.02 & 1.05 $\pm$ 0.03 & 1.03 $\pm$ 0.05 & 0.65 $\pm$ 0.01\\
\hline
conv5(array(1e6)) & 125.77 & 16.80 & 1.92 $\pm$ 0.03 & 1.09 $\pm$ 0.02 & 1.07 $\pm$ 0.05 & 0.80 $\pm$ 0.01\\
\hline
dilate3x3(Array2D(1000x1000)) & 23.29 & 12.99 & 0.41 $\pm$ 0.04 & 0.39 $\pm$ 0.01 & 0.26 $\pm$ 0.00 & 0.26 $\pm$ 0.01\\
\hline
sobel(Array2D(1000x1000)) & - & - & - & - & 0.19 $\pm$ 0.01 & 0.20 $\pm$ 0.01\\
\hline
%sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 $\pm$ 0.01 & 0.51 $\pm$ 0.00 & - & -\\
%\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93 $\pm$ 0.11 & 2.14 $\pm$ 0.02 & 0.97 $\pm$ 0.02 & 0.96 $\pm$ 0.01\\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 $\pm$ 0.03 & 1.00 $\pm$ 0.00 & 0.98 $\pm$ 0.00 & 0.98 $\pm$ 0.00\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 $\pm$ 0.01 & 1.82 $\pm$ 0.01 & 0.81 $\pm$ 0.01 & 0.80 $\pm$ 0.01\\
\hline

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class,  Array2D.
It is
a simple straight forward implementation providing 2 dimensionall
indexing with out of bounds checks.

@@

The current approach still has some limitations which we plan to address in the
future. In particular loop peeling works poorly in combination with trace
trees or trace stitching. The side exits attached guards that fail often
currently have to jump to the preamble which makes loops with several equally
common paths less efficient than they could be.
